### 2009-05-26 paper.tex
added SimulFold ref

We implemented these algorithms for a three-taxon phylogeny.  Using
our implementation on a simple model of pairwise RNA structural
evolution, we simulated data under the model and tested the dependence
of alignment inference and ancestral structure reconstruction accuracy on the evolutionary
distance of the outgroup of the phylogeny.  Our results suggest that
we may be able to accurately reconstruct ancestral structures despite
sequence and structural divergence if the evolutionary process is
well-understood.
We discuss the implications of our results for RNA algorithm design
and phylogenetic reconstruction of RNA evolutionary history.

@@

% (second part: multiple)
The second half of the problem involves extending the model (and related inference algorithms) from a single branch to a complete phylogeny,
i.e., from a pairwise model of two sequences to a {\bf multiple-sequence model} of many sequences.
In a typical situation, the sequences at the leaves of the tree are observed, but those at internal nodes are unobserved.
Questions of interest then include:
\begin{description}
\item[A.] What is the likelihood for the observed sequence data?
\item[B.] Can we sample from (or exactly compute) the posterior distribution of the unobserved sequence at the root node?
\item[C.] Can we sample from the posterior of the unobserved sequences at the other internal nodes?
\item[D.] Can we estimate summaries of the evolutionary history, such as
  the number of substitution events on each branch (for a substitution
  model), the alignment (for a model which includes indels), or
  changes in the underlying structure (for a model of RNA structure)?
\end{description}

@@

Reconstructing structural changes of large RNAs
using the three-way sampling kernel which we have described
would require resources far in excess of those currently available;
barring the availability of supercomputers with terabytes of memory,
such algorithms will only be feasible for short RNAs (\tabref{complexities}).
Thus, alternate strategies will have to be explored, such as the
progressive reconstruction used by the \ortheus\ program for
reconstructing ancestral genomes \cite{PatenEtAl2008}.
Another promising direction is to consider variations on the three-way sampling kernel,
such as the importance-sampling approach described for the TKF model
by \cite{RedelingsSuchard2005}.
This approach first proposes an ancestor $W$ by aligning extant
sequence $X$ to $Y$ (ignoring $Z$); then, in a second step, the proposed $W$ is
independently aligned to $Z$. The proposed three-way alignment and
reconstruction is then randomly accepted (or rejected) using a
Hastings ratio based on the three-way transducer composition. The
complexity of this kernel is the same as the pairwise case;
with suitable constraints, this is feasible for RNA
grammars on present hardware, at least for ribosomal domains (if not
yet whole subunits---although pairwise alignment of those should also
be possible soon). The approach of Redelings and Suchard therefore
merits future consideration in the context of modeling the evolution
of RNAs on a tree.

The evolutionary models we have described form a theoretical
foundation for studying the structural evolution of RNA gene families
in detail. By opening the door to ancestral sequence reconstruction
and paleogenomics, these algorithms will allow us to test our
knowledge of RNA evolution by direct experimental investigation
of early ribonucleic machines.

>>>>

We implemented these algorithms for a three-taxon phylogeny.  Using
our implementation on a simple model of pairwise RNA structural
evolution, we simulated data under the model and tested the dependence
of alignment inference and ancestral structure reconstruction accuracy on phylogenetic branch length.
Our results suggest that
we may be able to accurately reconstruct ancestral structures despite
sequence and structural divergence if the evolutionary process is
well-understood.
We discuss the implications of our results for RNA algorithm design
and phylogenetic reconstruction of RNA evolutionary history.

@@

% (second part: multiple)
The second half of the problem involves extending the model (and related inference algorithms) from a single branch to a complete phylogeny,
i.e., from a pairwise model of two sequences to a {\bf multiple-sequence model} of many sequences.
In a typical situation, the sequences at the leaves of the tree are observed, but those at internal nodes are unobserved.
Questions of interest then include:
\begin{description}
\item[A.] What is the likelihood for the observed sequence data?
\item[B.] Can we sample (find the mode, take moments, etc.) from the posterior distribution of the unobserved sequence at the root node?
\item[C.] Can we sample (etc.) from the posterior of the unobserved sequences at the other internal nodes?
\item[D.] Can we estimate summaries of the evolutionary history, such as
  the number of substitution events on each branch (for a substitution
  model), the alignment (for a model which includes indels), or
  changes in the underlying structure (for a model of RNA structure)?
\end{description}

@@

Reconstructing structural changes of large RNAs
using the three-way sampling kernel which we have described
would require resources far in excess of those currently available;
barring the availability of supercomputers with terabytes of memory,
such algorithms will only be feasible for short RNAs (\tabref{complexities}).
Thus, alternate strategies will have to be explored, such as the
progressive reconstruction used by the \ortheus\ program for
reconstructing ancestral genomes \cite{PatenHolmesBirney2008}.
Another promising direction is to consider variations on the three-way sampling kernel,
such as the importance-sampling approach described for the TKF model
by \cite{RedelingsSuchard2005}.
This approach first proposes an ancestor $W$ by aligning extant
sequence $X$ to $Y$ (ignoring $Z$); then, in a second step, the proposed $W$ is
independently aligned to $Z$. The proposed three-way alignment and
reconstruction is then randomly accepted (or rejected) using a
Hastings ratio based on the three-way transducer composition. The
complexity of this kernel is the same as the pairwise case;
with suitable constraints, this is feasible for RNA
grammars on present hardware, at least for ribosomal domains (if not
yet whole subunits---although pairwise alignment of those should also
be possible soon). The approach of Redelings and Suchard therefore
merits future consideration in the context of modeling the evolution
of RNAs on a tree.

An alternative MCMC scheme for sampling RNA phylogeny, structure and alignment was developed for the SimulFold program \cite{MeyerMiklos2007}.
SimulFold does not use a strictly normalized probabilistic model, resulting in some oddities in the ways that structure and indels interact
(for example, it does not penalize deletion of one half of a basepair).
Currently, it is not clear how appropriate SimulFold would be for ancestral reconstruction,
although it has several advantages (e.g. explicit treatment of pseudoknots).
Of course, MCMC kernels can often be adapted for use with different models, and the MCMC moves used by SimulFold may be useful for inference under our model as well.

In summary, the evolutionary models we have described form a theoretical
foundation for studying the structural evolution of RNA gene families
in detail. By opening the door to ancestral sequence reconstruction
and paleogenomics, these algorithms will allow us to test our
knowledge of RNA evolution by direct experimental investigation
of early ribonucleic machines.
